---
title: "KMeans"
output: html_document
date: "2026-05-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(plotly)
library(cluster)
library(factoextra)
library(corrplot)
library(ggplot2)
library(GGally)
library(imager)
library(animation)
library(data.table)
library(caret)
```

# Load Data

```{r}
dt <- fread("ecg_data.csv")

# Remove observations without labels
dt <- dt[!is.na(label)]
```

# Feature Engineering

The original HRV features were extended with lag, rolling statistics and delta features to capture temporal information.

```{r}
dt[, `:=`(
  lag1_feat1 = shift(vlf, 1),
  lag2_feat1 = shift(vlf, 2),

  roll3_mean = frollmean(vlf, 3, align = "right"),
  roll3_sd   = frollapply(vlf, 3, sd, align = "right"),

  roll5_mean = frollmean(vlf, 5, align = "right"),
  roll5_sd   = frollapply(vlf, 5, sd, align = "right"),

  delta1_feat1 = vlf - shift(vlf, 1),
  delta5_feat1 = vlf - shift(vlf, 5),

  lag1_feat2 = shift(rr_sd, 1),
  lag2_feat2 = shift(rr_sd, 2),

  roll3_mean_2 = frollmean(rr_sd, 3, align = "right"),
  roll3_sd_2   = frollapply(rr_sd, 3, sd, align = "right"),

  roll5_mean_2 = frollmean(rr_sd, 5, align = "right"),
  roll5_sd_2   = frollapply(rr_sd, 5, sd, align = "right"),

  delta1_feat2 = rr_sd - shift(rr_sd, 1),
  delta5_feat2 = rr_sd - shift(rr_sd, 5),

  lag1_feat3 = shift(bpm, 1),
  lag2_feat3 = shift(bpm, 2),

  roll3_mean_3 = frollmean(bpm, 3, align = "right"),
  roll3_sd_3   = frollapply(bpm, 3, sd, align = "right"),

  roll5_mean_3 = frollmean(bpm, 5, align = "right"),
  roll5_sd_3   = frollapply(bpm, 5, sd, align = "right"),

  delta1_feat3 = bpm - shift(bpm, 1),
  delta5_feat3 = bpm - shift(bpm, 5),

  lag1_feat4 = shift(hf, 1),
  lag2_feat4 = shift(hf, 2),

  roll3_mean_4 = frollmean(hf, 3, align = "right"),
  roll3_sd_4   = frollapply(hf, 3, sd, align = "right"),

  roll5_mean_4 = frollmean(hf, 5, align = "right"),
  roll5_sd_4   = frollapply(hf, 5, sd, align = "right"),

  delta1_feat4 = hf - shift(hf, 1),
  delta5_feat4 = hf - shift(hf, 5)

), by = patient]

dt <- na.omit(dt)
```

# Train Test Split

```{r}
train_patients <- c(
  "a01","a03","a04","a05","a06","a07","a08",
  "a09","a10","a11","a12","a13","a14","a15",
  "a16","b01","c03","c05","c06"
)

test_patients <- c(
  "a17","a18","a19","a20","b03","c07","c08"
)

training <- dt[patient %in% train_patients]
test <- dt[patient %in% test_patients]
```

# Feature Selection

Use all engineered features except metadata.

```{r}
#feature_cols <- c( "rr_mean","rr_sd","rmssd","nn50", "pnn50","bpm","vlf","lf", "hf", "lf_hf","lf_norm","hf_norm")
feature_cols <- setdiff(
  names(training),
  c("label", "patient", "title")
)

train_x <- as.data.frame(training[, ..feature_cols])
train_y <- factor(training$label)

test_x <- as.data.frame(test[, ..feature_cols])
test_y <- factor(test$label)
```

# Feature Scaling

K-Means is distance-based and therefore requires standardized features.

```{r}
train_scaled <- scale(train_x)

center <- attr(train_scaled, "scaled:center")
scale_ <- attr(train_scaled, "scaled:scale")

train_x_scaled <- as.data.table(train_scaled)

test_x_scaled <- as.data.table(
  scale(test_x, center = center, scale = scale_)
)
```
pca test
```{r}
#pca <- prcomp(train_x_scaled,center = TRUE,scale. = TRUE)

#summary(pca)#summary(pca)$importance#train_pca <- pca$x[,1:15]

#test_pca <- predict(pca,newdata = test_x_scaled)[,1:15]
```

# Finding the Best k

The elbow method and silhouette score were used to evaluate different cluster numbers.

```{r}
set.seed(123)

k_values <- 2:10

results <- data.frame(
  k = integer(),
  WSS = numeric(),
  Silhouette = numeric()
)

for(current_k in k_values){

  km <- kmeans(
    train_x_scaled,
    centers = current_k,
    nstart = 50,
    iter.max = 100
  )

  sil <- silhouette(
    km$cluster,
    dist(train_x_scaled)
  )

  results <- rbind(
    results,
    data.frame(
      k = current_k,
      WSS = km$tot.withinss,
      Silhouette = mean(sil[,3])
    )
  )
}
```

```{r}
print(results)
```
Plotting the results
```{r}
plot_ly(type = "scatter", mode = "lines") |>
  add_lines(
    x = results$k,
    y = results$WSS,
    name = "WSS"
  ) |>
  add_lines(
    x = results$k,
    y = results$Silhouette,
    name = "Silhouette"
  )
```

# Final Model

Based on the previous experiments, the final model uses four clusters.

```{r}
set.seed(123)

final_k <- 4

kmeans_model <- kmeans(
  train_x_scaled,
  centers = final_k,
  nstart = 50,
  iter.max = 100
)
```

# Assign Cluster Labels

K-Means returns cluster IDs. Therefore, clusters are mapped to the majority class in the training data.

```{r}
cluster_label_map <- data.table(
  cluster = kmeans_model$cluster,
  label = train_y
)

cluster_counts <- cluster_label_map[
  , .N,
  by = .(cluster, label)
]

mapping <- cluster_counts[
  order(cluster, -N)
][
  , .SD[1],
  by = cluster
]

mapping
```

# Predict Training Data

```{r}
train_clusters <- kmeans_model$cluster

train_pred <- sapply(
  train_clusters,
  function(x){
    mapping[cluster == x, as.character(label)]
  }
)

train_pred <- factor(
  train_pred,
  levels = c("0","1")
)
```

# Predict Test Data

```{r}
centers <- kmeans_model$centers

predict_kmeans <- function(newdata, centers){

  distances <- sapply(
    1:nrow(centers),
    function(i){
      rowSums(
        (sweep(newdata, 2, centers[i,], "-"))^2
      )
    }
  )

  max.col(-distances)
}

test_clusters <- predict_kmeans(
  as.matrix(test_x_scaled),
  centers
)

test_pred <- sapply(
  test_clusters,
  function(x){
    mapping[cluster == x, as.character(label)]
  }
)

test_pred <- factor(
  test_pred,
  levels = c("0","1")
)
```

# Evaluation

## Test Set

```{r}
kmeans_confusion_matrix <- confusionMatrix(
  test_pred,
  test_y,
  positive = "1",
  mode = "prec_recall"
)

print(kmeans_confusion_matrix)
```

## Training Set

```{r}
kmeans_confusion_matrix_train <- confusionMatrix(
  train_pred,
  train_y,
  positive = "1",
  mode = "prec_recall"
)

print(kmeans_confusion_matrix_train)
```

# Extract Metrics

```{r}
data.frame(
  Accuracy = kmeans_confusion_matrix$overall["Accuracy"],
  Recall = kmeans_confusion_matrix$byClass["Recall"],
  Precision = kmeans_confusion_matrix$byClass["Precision"],
  F1 = kmeans_confusion_matrix$byClass["F1"]
)
```


```{r}
## Per Patient Accuracy

patient_perf <- test_results[, .(
  Accuracy = mean(pred == as.character(label)),
  N = .N,
  Apnea_Rate = mean(label == "1")
), by = patient]

print(patient_perf)
```